The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf

### 2015-03-31 oil_prices_rr.tex
update on paper

% revised paper for Energy Journal

@@

% table font size
\let\oldtabular\tabular
\renewcommand{\tabular}{\scriptsize\oldtabular}

@@

\begin{abstract}
I use detailed field-level data on Norwegian off-shore oil field production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.  I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 4\% for a 10 dollar per barrel increase in the real price of oil.  Most of this effect appears to come in the planning phase of a field's development.\\
Keywords: Oil production,Oil Prices, Norway, Semiparametric.
\end{abstract}

@@

For most of the last century, crude oil has been the world’s single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013 as well as the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}} Naturally, questions of how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics in economics.

@@

The main finding in this article is that oil production at the field level has no significant reaction to concurrent changes in the oil price, where concurrent is broadly defined as within the first three years.  A slight effect can be detected at a lag of between 4 and 8 years, with a magnitude of about a 2 to 4\% increase in yearly production for a 10 dollar increase in the price of oil.  This effect is somewhat greater and with less of a lag in large fields compared to small fields.  Price appears to have the most significant affect during the planning stage - before production begins in a field.  In the depletion phase of production, price is found to have little to no significant effect.

@@

More so, oil field production is correlated across fields - that is, increases and decreases in production in fields are not randomly distributed across time.  Instead, as figure \ref{top10_production} shows with the production profile of the 10 largest Norwegian oil fields, production tends to be correlated across fields.  The result is a total production curve that is bell-shaped over time as in figure \ref{oil_decline}.  Since the oil price series is autocorrelated and non-stationary, failure to properly account for the production profile will lead to spurious estimation of the price terms in a regression.

@@

As a solution I use a semi-parametric model within the Generalized Additive Model frameworks of \cite{hastie_generalized_1990}.  Here I use a cubic spline function to account for the general non-parametric shape of the production profile while allowing price to enter the equation linearly while also attempting to control for the effects of varying production costs and the effects of technological change and other time-varying effects.

The results suggest that little to no concurrent effect of oil prices on production exists.  However modest lagged effects of between 2 to 5 years exists.  These results are in line with the intuition

@@

For modeling aggregate oil production, shape-fitting models, notably \citet{hubbert_energy_1962} and more recently advocated by \citet{deffeyes_hubberts_2001}, have had some success in estimating the timing of peak production at the regional and national level, but they tend to seriously underestimate the total recoverable resource of oil-producing regions and the models can be shown to be fundamentally misspecified \citep{boyce_prediction_2013}.   Simulation type studies where aggregate oil production is modeled through an often complex combination of physical and economic processes also exist in the geo-engineering literature, but their usefulness tends to be weighed down by their complexity as they require quite detailed data and specific assumptions about functional form that can be difficult to justify \citet{brandt_review_2010}.

@@

the effect of oil-price uncertainty on drilling and exploration has also been explored.  \citet{kellogg_effect_2014} finds that oil exploration firms in Texas do approximately respond as real-option theory would predict when it comes to the timing of drilling.  A model and test using data from North Sea producers on the UK continental shelf by \citet{hurn_geology_1994}, on the other hand, fails to find evidence that the variance in the oil price affects the timing of oil field development.  I do not attempt to directly model uncertainty, however given that the investments needed to increase oil production in an existing field are to a certain extent irreversible and that oil prices are highly volatile, the results can and probably should be interpreted with the real options framework in mind.

Only a few papers utilize field-level data.  \citet{black_is_1998} tests the relevance of “nesting” a structural empirical model of profit maximization that takes into account oil prices into a typical geo-engineering model of oil field production.  They find strong evidence that taking into account profit maximization, and implicitly price, substantially improves the fit compared to a purely geo-engineering type model.   The limitation of their methodology is that they are only able to test whether including economic factors like price affects the fit of the model but are not able to give an estimate of the effect.  Methodologically, the paper also relies heavily on assumptions about the functional form of both the geo-engineering aspects of the oil producer as well as their profit-maximization.  By taking a more flexible, semi-parametric approach to estimating the effects of oil field production profile, this paper avoids problems with overly restrictive assumptions.

\citet{rao_taxation_2010} uses data on land based oil wells in California to estimate the effects of tax changes and price controls.  The author finds that short-term tax changes caused small, but significant retiming of production from oil wells.  These findings are over-all consistent with the findings of this paper, though the author finds concurrent effects of taxation on production while I do not find any concurrent effect of prices on production.  This difference is most likely due to the significant differences in cost and complexity between operating offshore and onshore.  How producers react to a short-term tax change as opposed to a change in price is likely also different.  More so, the author does not consider differences between the different stages of production.

Studies using detailed Norwegian data on offshore activity also exist, though the focus has mainly been on exploration and drilling.  \citet{mohn_exploration_2008} finds that long-term changes in the oil-price has a strong effect on exploratory drilling though little effect is measured from short-term changes in the oil price.  \citet{osmundsen_exploration_2010} analyses drilling productivity over time on the Norwegian Continental Shelf while \citet{mohn_efforts_2008} finds that higher oil prices leads to higher reserves and as well as that oil prices affect producer risk-preferences - with higher prices leading to lower success rates but larger discovery size.

@@

The first commercial oil well in Norwegian continental waters was discovered in December of 1969 in what became the Ekofisk oil field, the largest Norwegian oil field by estimated recoverable reserves.  As figure \ref{north_sea_reserves} shows, most of the largest fields in the North Sea were found relatively early on while more recent finds have tended to be smaller - a pattern typical of oil producing areas called creaming.  A major exception to this trend was the recent find of the Johan Sverdrup field which is estimated to have approximately 300 million Standard Cubic Meters (SM3) of recoverable oil. \footnote{The Johan Sverdrup field is estimated to begin producing oil in 2019 and so is not present in the data set used for the analysis.}

Exploration in the Norwegian Sea was opened in the early 1980’s and the first commercial field started production in 1981.  While several mid-sized fields have been discovered, the Norwegian Sea has generally disappointed in terms of commercial oil finds and most finds have been relatively small (\ref{norwegian_sea_reserves}).

Norwegian waters in the Barents Sea have also been open to exploration since the 1980s, however up until recently only a few, small finds were made and none came into commercial production.  However recently several significant oil and gas finds have been made in the Barents Sea - notably the oil field Goliat and the large gas find Sn\o hvit, which are both currently under development but not yet producing.  The agreement between Norway and Russia in June 2011 settling a long-running dispute over the maritime delimitation has also given a boost to new exploration in the region.

Profits from oil and gas production in Norway are subject to a resource tax of 50 \% on top of the ordinary income tax of 28 \%, thus income from petroleum production is taxed at a total marginal tax rate of 78 \%.  The central government also receives revenues through ownership stakes in companies, notably Statoil, where the state is the majority stakeholder.  The over-all tax rate has been fairly constant through the history of Norwegian oil production, however several important changes related to the tax code have occurred.  In 1991 a C02 tax was introduced, and in the year 2012 the tax was doubled.  But this was mainly levied on the use and import of petroleum products.  In the offshore sector it was levied on the burning of oil and gas and thus the main effect was on the practice of flaring natural gas that could not be transported and sold commercially.

More importantly to the offshore sector were accounting changes that were implemented in 2002 and 2005 that were meant to encourage new entrants by allowing losses to be carried forward for tax purposes and by introducing a rebate on the tax-value of losses associated with searching and drilling.  In general though, these rules mainly affected searching and discoveries of new fields rather than production from existing fields and I do not control for the tax changes in my model.  More information on taxation and revenues from the offshore sector can be found at the website of the Norwegian Ministry of Finance. \footnote{\url{http://www.regjeringen.no/en/dep/fin/Selected-topics/taxes-and-duties/bedriftsbeskatning/Taxation-of-petroleum-activities.html?id=417318}}

@@

\section{Norwegian oil field production data}
Production data of Norwegian oil fields is obtained from the website of the Norwegian Petroleum Directorate. \footnote{http://factpages.npd.no/}  Production data is available at a monthly frequency for all fields, though I choose to aggregate up to yearly values both to smooth over seasonality as well as short-term volatility of output due to factors such as weather or technical issues that are not relevant for this article.

In addition to data on field-level production, I also make use of data on estimated total recoverable reserves. The use of this variable is complicated as it is an estimate subject to a large amount of uncertainty, especially in relatively young fields.  However, the methodologies used to estimate the total recoverable resource of a field are constantly evolving and it is a fair assumption that any consistent bias of the estimates are observed in older fields and corrected for in estimates for newer fields.  I can then assume that existing errors are random and will not significantly bias the estimates.

Moreover, the estimate is likely endogenous, in the sense that it is also effected by prices.  However since I use the variable as a control variable and not for the purpose of estimating a parameter with a causal interpretation, this should not significantly affect the validity of the results.

I use yearly data from the US Energy Information Agency on the real price of Brent-traded oil in 2010 dollars.The Brent benchmark oil price is likely the best oil price measure for Norwegian production as it is based on light sweet crude oil sourced from the North Sea.

@@

On the other hand, given the size and liquidity of oil spot markets, it is a fair assumption that the current oil prices do a good job of incorporating much of the available information about crude oil markets and that future price movements are generally difficult to predict \citep{hamilton_understanding_2008}.  An active futures market does exist, but several studies have found that current oil prices are in general better than prices on futures contracts at predicting future oil prices \citep{alquist_what_2010, chinn_predictive_2005}.  \citet{mohn_investment_2008} as well as \citet{pesaran_econometric_1990} and \citet{farzin_impact_2001} find evidence for adaptive expectations, where expectations of future prices is based on a weighted average of current and past prices.  I take account of this by including several years of price lags in my regression equations.

A cleaned data set and the full code for the analysis are available at my website \url{jmaurit.github.io#oil_prices}.

\section{A generalized additive model of oil field production}
Parametric linear models have the sizeable advantages of simplicity and interpretability and are therefore usually a good starting point for an analysis.  However, when attempting to model the effect of price on oil field production, a standard linear model is unable to sufficiently control for the production profile and therefore heavily biases the estimate of the effect of price.  As an example, consider a generalized linear model written as in equation \ref{glm_eqn}.

	\begin{equation}
	\begin{split}	
	 Log(Production_{i,t})  &= \alpha_0 + \alpha_1 timetopeak_{i,t} + \alpha_2 timetopeak_{i,t}^2 \\
	  \quad & + \alpha_3 timetopeak_{i,t}^3  + \alpha_4 peaktoend_{i,t} + \alpha_5 peaktoend_{i,t}^2 \\
	  \quad & + \alpha_6 peaktoend_{i,t}^3 + \gamma totalrecoverableoil_i \\
	  \quad & + \beta_1 oilprice + \beta_2 oilpricel1 + ...+ \beta_6 oilpricel5 + \epsilon	
	\label{glm_eqn}
	\end{split}
	\end{equation}

This estimate is heavily biased downwards due to the spurious correlation between the field production profiles and the autocorrelated and non-stationary time series of oil prices.  A more formal analysis of the bias can be found in Appendix B, where I simulate data of oil field production in a Monte Carlo experiment.  However, the basic idea is that the parametric representation is not flexible enough to control for the production profiles of the fields.   Instead, a more flexible estimation of the production profile is needed.

Instead of attempting to estimate the shape of the production profiles of the fields by estimating parameters on linear terms I estimate a non-parametric function for the production profile.  As an illustration, consider the production profile of a single field.  The simplest possible model would then have the form of equation \ref{simp_eqn}.

In principle any number of well-behaved smoothers could be used to estimate the function, for example a Loess or a kernel smoother.  In practice regression splines are most commonly used.  With a regression spline the data points for the function to be estimated are broken into bins.  For each bin of data a local linear regression is estimated.  For functions of one variable, a cubic parameterization is often used.  These regressions are then essentially tied together at what are called “knots” and the smoothness of the overall function is controlled by a penalty function that consists of the second derivative of the estimated function.

The advantage to using splines over other smoothing methods is that it can be represented in a linear form.  Thus estimation of the model can be done using standard and efficient matrix algebra algorithms. For further details I refer to the discussion in \citet{hastie_generalized_1990} and \citet{wood_generalized_2006}.  The latter is a particularly useful reference for implementing generalized additive models in R.

I do not want to estimate smoothed curves individually for each field.  While this would provide a good overall fit to the full data set, not enough variation in the data would be left to estimate the effect of price.  Instead I want to estimate a general shape of the production profile for all fields and then use the remaining variation in the data to estimate the effect of price.  My model can be written as in equations \ref{gam_price_eqn}.

@@

	Log(Production_{i,t}) & = f(time\_to\_peak_{i,t}, total\_recoverable\_oil_i) \\
 	 \quad & + f(peak\_to\_end_{i,t}, total\_recoverable\_oil_i) \\
	 \quad & + \beta_1 oil\_price + \beta_2 oil\_price\_l1 + ... +  \epsilon
\label{gam_price_eqn}
\end{split}
\end{equation}

In this model I am estimating the parameters and functions from all fields $i$. As in the parametric model presented earlier, the left-hand-side variable is yearly oil production for field $i$. Also like the parametric model I split the production-time component element in two: up to and after the peak in production.

@@

\begin{table}
\begin{center}
\begin{tabular}{l c c c }
\hline
                                 & w/out Rand. Effects & w Rand. Effects & Lags 1-5 \\
\hline
(Intercept)                      & $0.335$        & $-4.016^{***}$ & $-3.971^{***}$ \\
                                 & $(0.265)$      & $(0.297)$      & $(0.212)$      \\
I(in\_place\_oil\_mill\_sm3/100) & $0.337^{***}$  & $-0.014$       & $-0.015$       \\
                                 & $(0.009)$      & $(0.022)$      & $(0.022)$      \\
I(cost\_index/10)                & $-0.018^{**}$  & $-0.030^{***}$ & $-0.030^{***}$ \\
                                 & $(0.006)$      & $(0.006)$      & $(0.005)$      \\
I(price/10)                      & $-0.006$       & $0.014$        &                \\
                                 & $(0.017)$      & $(0.016)$      &                \\
I(price\_l1/10)                  & $0.004$        & $0.021$        & $0.029^{*}$    \\
                                 & $(0.017)$      & $(0.017)$      & $(0.014)$      \\
I(price\_l2/10)                  & $0.037^{*}$    & $0.047^{**}$   & $0.048^{***}$  \\
                                 & $(0.017)$      & $(0.016)$      & $(0.014)$      \\
I(price\_l3)                     & $0.003$        & $0.003$        & $0.002$        \\
                                 & $(0.002)$      & $(0.001)$      & $(0.001)$      \\
I(price\_l4)                     & $0.000$        & $0.004^{**}$   & $0.004^{***}$  \\
                                 & $(0.002)$      & $(0.001)$      & $(0.001)$      \\
I(price\_l5/10)                  & $-0.020$       & $0.012$        & $0.019$        \\
                                 & $(0.013)$      & $(0.012)$      & $(0.010)$      \\
I(price\_l6/10)                  & $-0.010$       & $0.008$        &                \\
                                 & $(0.012)$      & $(0.011)$      &                \\
I(price\_l7/10)                  & $-0.008$       & $0.007$        &                \\
                                 & $(0.012)$      & $(0.012)$      &                \\
I(price\_l8/10)                  & $-0.044^{***}$ & $-0.011$       &                \\
                                 & $(0.011)$      & $(0.011)$      &                \\
EDF: s(prod\_year)               & $6.381^{***}$  &                &                \\
                                 & $(7.249)$      &                &                \\
EDF: s(year)                     & $2.768^{***}$  & $3.000^{***}$  & $3.000^{***}$  \\
                                 & $(2.956)$      & $(3.000)$      & $(3.000)$      \\
EDF: s(prod\_year,name)          &                & $33.002^{***}$ & $33.014^{***}$ \\
                                 &                & $(72.000)$     & $(72.000)$     \\
\hline
AIC                              &                &                &                \\
BIC                              &                &                &                \\
Log Likelihood                   &                &                &                \\
Deviance                         & 2306267.814    & 1976585.232    & 1980468.734    \\
Deviance explained               & 0.848          & 0.870          & 0.870          \\
Dispersion                       & 1785.243       & 1562.520       & 1560.671       \\
R$^2$                            & 0.841          & 0.865          & 0.866          \\
GCV score                        & 1814.469       & 1621.811       & 1614.802       \\
Num. obs.                        & 1088           & 1088           & 1088           \\
Num. smooth terms                & 2              & 2              & 2              \\
\hline
\multicolumn{4}{l}{\scriptsize{$^{***}p<0.001$, $^{**}p<0.01$, $^*p<0.05$}}
\end{tabular}
\caption{Statistical models}
\label{GAM_model_table}
\end{center}
\end{table}

>>>>

@@

@@

\begin{abstract}
I use detailed field-level data on Norwegian off-shore oil field production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.  I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 3 to 5\% for a 10 dollar per barrel increase in the real price of oil.  Most of this effect appears to come in the build-out or planning phase of a field's development.\\
Keywords: Oil production,Oil Prices, Norway, Generalized Additive Model.
\end{abstract}

@@

For most of the last century, crude oil has been the world’s single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013 as well as the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}} Naturally, questions of how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics in economics. Recent volatility in the price of oil has made this an especially relevant topic.

@@

The main finding in this article is that oil production at the field level has no significant reaction to concurrent changes in the oil price.  A slight effect can be detected at a lag of between 1 and 4 years, with a magnitude of about a 2 to 5\% increase in yearly production for a 10 dollar increase in the price of oil.

% Price appears to have the most significant affect during the planning and build-out stage.  In the depletion phase of production, price is found to have little to no significant effect.

@@

More so, oil field production is correlated across fields - that is, increases and decreases in production in fields are not randomly distributed across time.  Instead, as the top panel of figure \ref{oil_decline} shows with the production profile of the 10 largest Norwegian oil fields, production tends to be correlated across fields.  The result is a total production curve that is bell-shaped over time as in the lower panel of figure \ref{oil_decline}.  Since the oil price series is autocorrelated and non-stationary, failure to properly account for the production profile will lead to spurious estimation of the price terms in a regression.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/oil_decline.png}
	\caption{The panel shows the production from the largest 10 fields, which are temporally correlated with each other.  The total production over time is bell-shaped, as shown by the bottom panel.}
	\label{oil_decline}
\end{figure}

@@

As a solution I use a semi-parametric model within the Generalized Additive Model frameworks of \cite{hastie_generalized_1990}.  Here I use both a cubic spline and a two-dimensional thin-plate spline to account for the general non-parametric shape of the production profile and allowing price to enter the equation linearly while also attempting to control for the effects of varying production costs and the effects of technological change and other time-varying effects.

The results suggest that little to no concurrent effect of oil prices on production exists.  However modest lagged effects of between 1 to 4 years exist.These results are in line with an understanding of the industry as being highly capital intensive with high operating costs.  Maintaining spare capacity in order to adjust to changing oil prices is likely prohibitively inexpensive.  Instead, production is increased through investment, which may take up to several years to show an effect.

@@

For modeling aggregate oil production, shape-fitting models, notably \citet{hubbert_energy_1962} and more recently advocated by \citet{deffeyes_hubberts_2001}, have had some success in estimating the timing of peak production at the regional and national level, but they tend to seriously underestimate the total recoverable resource of oil-producing regions and the models can be shown to be fundamentally misspecified \citep{boyce_prediction_2013}.   Simulation type studies where aggregate oil production is modeled through an often complex combination of physical and economic processes also exist in the geo-engineering literature, but their usefulness tends to be weighed down by their complexity as they require quite detailed data and specific assumptions about functional form that can be difficult to justify \citep{brandt_review_2010}.

@@

the effect of oil-price uncertainty on drilling and exploration has also been explored.  \citet{kellogg_effect_2014} finds that oil exploration firms in Texas do approximately respond as real-option theory would predict when it comes to the timing of drilling.  A model and test using data from North Sea producers on the UK continental shelf by \citet{hurn_geology_1994}, on the other hand, fails to find evidence that the variance in the oil price affects the timing of oil field development.

I do not attempt to directly model uncertainty, however given that the investments needed to increase oil production in an existing field are to a certain extent irreversible and that oil prices are highly volatile, the results can and probably should be interpreted with the real options framework in mind.

Only a few papers utilize field-level data.  \citet{black_is_1998} tests the relevance of “nesting” a structural empirical model of profit maximization that takes into account oil prices into a typical geo-engineering model of oil field production.  They find strong evidence that taking into account profit maximization - and implicitly price - substantially improves the fit compared to a purely geo-engineering type model.   The limitation of their methodology is that they are only able to test whether including economic factors like price affects the fit of the model but are not able to give an estimate of the effect.  Methodologically, the paper also relies heavily on assumptions about the functional form of both the geo-engineering aspects of the oil producer as well as their profit-maximization.  By taking a more flexible, semi-parametric approach to estimating the effects of oil field production profile, this paper avoids problems with overly restrictive assumptions.

\citet{rao_taxation_2010} uses data on land-based oil wells in California to estimate the effects of tax changes and price controls.  The author finds that short-term tax changes caused small, but significant retiming of production from oil wells.  These findings are over-all consistent with the findings of this paper, though the author finds concurrent effects of taxation on production while I do not find any concurrent effect of prices on production. The significant differences in cost and complexity between operating offshore and onshore likely account for the different results.  How producers react to a short-term tax change as opposed to a change in price is likely also different.

Studies using detailed Norwegian data on offshore activity also exist, though the focus has mainly been on exploration and drilling.  \citet{mohn_exploration_2008} finds that long-term changes in the oil-price has a strong effect on exploratory drilling though little effect is measured from short-term changes in the oil price.  \citet{osmundsen_exploration_2010} analyses drilling productivity over time on the Norwegian Continental Shelf while \citet{mohn_efforts_2008} finds that higher oil prices leads to higher reserves as well as that oil prices affect producer risk-preferences - with higher prices leading to lower success rates but larger discovery size.

@@

The first commercial oil well in Norwegian continental waters was discovered in December of 1969 in what became the Ekofisk oil field, the largest Norwegian oil field by estimated recoverable reserves.  Most of the largest fields in the North Sea were found relatively early on while more recent finds have tended to be smaller - a pattern typical of oil producing areas called creaming.  A major exception to this trend was the recent find of the Johan Sverdrup field which is estimated to have approximately 300 million Standard Cubic Meters (SM3) of recoverable oil. \footnote{The Johan Sverdrup field is estimated to begin producing oil in 2019 and so is not present in the data set used for the analysis.}

Exploration in the Norwegian Sea was opened in the early 1980’s and the first commercial field started production in 1981.  While several mid-sized fields have been discovered, the Norwegian Sea has generally disappointed in terms of commercial oil finds and most finds have been relatively small.

Norwegian waters in the Barents Sea have also been open to exploration since the 1980s, however up until recently only a few, small finds were made and none came into commercial production.  Recently several significant oil and gas finds have been made in the Barents Sea - notably the oil field Goliat and the large gas find Sn\o hvit, which are both currently under development but not yet producing.  The agreement between Norway and Russia in June 2011 settling a long-running dispute over the maritime delimitation has also given a boost to new exploration in the region.

Profits from oil and gas production in Norway are subject to a resource tax of 51 \% on top of the ordinary income tax of 27 \%, thus income from petroleum production is taxed at a total marginal tax rate of 78 \%.  The central government also receives revenues through ownership stakes in companies, notably Statoil, where the state is the majority stakeholder.

The over-all tax rate has been fairly constant through the history of Norwegian oil production, however several important changes related to the tax code have occurred.  In 1991 a C02 tax was introduced, and in the year 2012 the tax was doubled.  But this was mainly levied on the use and import of petroleum products.  In the offshore sector it was levied on the burning of oil and gas and thus the main effect was on the practice of flaring natural gas that could not be transported and sold commercially.

More important to the offshore sector were accounting changes that were implemented in 2002 and 2005, which were meant to encourage new entrants by allowing losses to be carried forward for tax purposes and by introducing a rebate on the tax-value of losses associated with searching and drilling.  In general though, these rules mainly affected searching and discoveries of new fields rather than production from existing fields and I do not control for the tax changes in my model.  More information on taxation and revenues from the offshore sector can be found at the website of the Norwegian Ministry of Finance. \footnote{\url{http://www.regjeringen.no/en/dep/fin/Selected-topics/taxes-and-duties/bedriftsbeskatning/Taxation-of-petroleum-activities.html?id=417318}}

@@

\section{Data}
Production data of Norwegian oil fields is obtained from the website of the Norwegian Petroleum Directorate.\footnote{http://factpages.npd.no/} The data I use includes all current or formerly oil-producing fields, from 1971 through 2014. Production data is available at a monthly frequency, though I choose to aggregate up to yearly values both to smooth over seasonality as well as short-term volatility of output due to factors such as weather or technical issues that are not relevant for this article.

In addition to data on field-level production, I also have access to data on estimates of total recoverable reserves as well as estimates of total in-place oil for each field. The use of total recoverable reserves is complicated since it is dependent on the price of oil and thus endogenous in the analysis.  Therefor, as a proxy for the size of the fields, I use estimates of in-place oil from the Norwegian Petroleum Directorate, which is unlikely to be correlated with prices. The left panels of figure \ref{data_descriptives} show the histogram of field sizes in my data set.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/data_descriptives.png}
	\caption{Data descriptives. The left panels show histograms of estimates of recoverable oil and in-place oil by field. The right panels shows the production cost index and real brent oil prices.}
	\label{data_descriptives}
\end{figure}

To simplify the analysis, I eliminate the smallest fields from my analysis - those with less than 3 million Standard Cubic Meters (SM3) of recoverable reserves.  These fields tend to have substantially different production profiles than the larger fields. However, these fields only account for around .5 \% of total oil production so this exclusion is unlikely to have any serious practical implications.

I also eliminate the large Ekofisk field from my analysis.  I do this because it has a unique double-humped production profile. Ekofisk was the first field to be found and the first to start production on the Norwegian Continental Shelf.  It reached an initial peak in production in the late 1970's and began depleting. However, the introduction of water and gas injection technology led to a boost in production and a new peak in production in the early 2000's. All subsequent major fields used gas and water injection and have a single-hump profile.

I use yearly data from the US Energy Information Agency on the real price of Brent-traded oil in 2010 dollars. The Brent benchmark oil price is likely the best oil price measure for Norwegian production as it is based on light sweet crude oil sourced from the North Sea.

Production costs are known to vary substantially with the price of oil.  To control for this, I include the producer price index for Norwegian oil and gas extraction industry, obtained from Statistics Norway. \footnote{\url{https://www.ssb.no/en/priser-og-prisindekser/statistikker/ppi}} This series only goes back to 2000, however, so I also include an index of oil and gas well costs from the US energy information agency for the years 1960 to 2000.  \footnote{\url{http://www.eia.gov/dnav/pet/pet_crd_wellcost_s1_a.htm}} While this measure is unlikely to be a perfect substitute for costs in Norway, oil and gas extraction is a global industry, where costs are highly correlated across regions. The combined production cost index and the real Brent oil price series are shown in the right panels in figure \ref{data_descriptives}.

@@

On the other hand, given the size and liquidity of oil spot markets, it is a fair assumption that the current oil prices do a good job of incorporating much of the available information about crude oil markets and that future price movements are generally difficult to predict \citep{hamilton_understanding_2008}.

An active futures market does exist, but several studies have found that current oil prices are in general better than prices on futures contracts at predicting future oil prices \citep{alquist_what_2010, chinn_predictive_2005}.  \citet{mohn_investment_2008} as well as \citet{pesaran_econometric_1990} and \citet{farzin_impact_2001} find evidence for adaptive expectations, where expectations of future prices is based on a weighted average of current and past prices.  I take account of this by including several years of price lags in my regression equations.

A cleaned data set and the full code for the analysis are available at \url{jmaurit.github.io/#oil_prices}.

\section{A generalized additive model of oil field production}
Parametric linear models have the sizeable advantages of simplicity and interpretability and are therefore usually a good starting point for an analysis. However, when attempting to model the effect of price on oil field production, a standard linear model is unable to sufficiently control for the production profile and therefore heavily biases the estimate of the effect of price.

Instead of attempting to estimate the shape of the production profiles of the fields by estimating parameters on linear terms I estimate a non-parametric function for the production profile. At a conceptual level, this model is decomposing the variation in the data into a smoothed component representing the technical and physical forces that produce the typical bell-shaped production profile, and the remaining component, that I estimate the effect of price with.

I do not, however, want to estimate smoothed curves individually for each field. While this would provide a good overall fit to the full data set, not enough variation in the data would be left to estimate the effect of price.  Instead I want to estimate a general shape of the production profile for all fields and then use the remaining variation in the data to estimate the effect of price.

A simple model can be written as in equation \ref{gam_price_eqn}.

@@

	Log(production_{i,t}) & = f(production\_time_{i,t}) + \beta_1 in\_place\_oil_{i,t} \\
	 \quad & + \beta_2 oil\_price\_lags_t + \beta_3 cost\_index_{t} +  
	 f(year_t) + \epsilon_{i,t}
\label{gam_price_eqn}
\end{split}
\end{equation}

In this model I am estimating the parameters and functions from all fields $i$. The left-hand-side variable, $Log(production_{i,t})$ is the yearly oil production for field $i$ at production time $t$ in millions of standard cubic meters (SM3).  On the right hand side is a smoothed function of the production profile over time $f(production\_time_{i,t})$, $in\_place\_oil_{i,t}$ is the estimated in place oil of the field, and $cost\_index_{t}$ represents the producer price index for oil and gas extraction.

$f(year_t)$ represents a function of the calendar year.  The main purpose of including this term is to control for the effects of technological change.  I could use a smoothed function here as well, however in order to avoid over fitting the model, I use a simple quadratic representation.

Finally, $oil\_price\_lags_t$ represents the variables of interest - a vector of concurrent and up to 8 lagged terms for the oil price. I also experimented with using quadratic terms for the oil price and its lags, however these could not be shown to be significant.

The smoothed function is estimated using a cubic regression spline...

One objection to this model may be that a fixed effect controlling for the general production profile of oil fields does not sufficiently address idiosyncratic differences between the fields. To address this I can add the term, $\beta_r Z$, where $Z$ represents field random effects that takes into account uncertainty based on random variation between the fields.

Another substantial objection may be that the smoothed function of production time may not be sufficient in controlling for the variation in production profiles shapes across fields.  In particular, production profiles differ substantially by the size of the field. A solution is to allow the shape to vary by production.  Thus the smoothed can be written as a function of both production time and in place oil: $f(production\_time_{i,t}, in\_place\_oil_{i,t})$.

Estimating a smoothed function of two variables requires a somewhat more involved method.  In particular I use Thin-plate regression spline \citep{wood_thin_2003}. Consider equation \ref{thin_plate_1}.

	\begin{equation}
	y_i = g(x_1, x_2)
	\label{thin_plate_1}
	\end{equation}

Following the notation of \citet{wood_generalized_2006}, $g$ is the function of $x_1$ and $x_2$ that is to be estimated by $f$, which in turn is estimated by minimizing equation \ref{thin_plate_2}.  Here $\boldsymbol{y}$ represents a vector of $y_i$’s and $\boldsymbol{f} = (f(\boldsymbol{x_1}),f(\boldsymbol{x_2}))^t$.

	\begin{equation}
\min \|\boldsymbol{y-f}\|^2 + \lambda J_{22}(f)
\label{thin_plate_2}
	\end{equation}

$J_{22}$ represents the penalty function for the smoothness of the function which can be written as in \ref{thin_plate_3}.  The $22$ represents the fact that it is a penalty function of two variables with smoothness measured by the second derivative.

	\begin{equation}
	J_{22}{f}= \diffp[2]{f}{x_1}^2 + \diffp[2]{f}{{x_1}{x_2}} + \diffp[2]{f}{x_2}^2dx_1 dx_2
\label{thin_plate_3}
	\end{equation}

In short, a function of $x_1$ and $x_2$ is found that is minimizing errors in the sense of minimizing Euclidean distance subject to a penalty function of “wiggiliness”.  The actual implementation is somewhat more involved in order to increase the computational efficiency of the estimation.  For further details I again refer to \citet{wood_thin_2003}.

@@

\begin{table}
\begin{center}
\begin{tabular}{l c c c }
\hline
                                             & w/out Rand. Effects & w Rand. Effects & 2-d Smooth \\
\hline
(Intercept)                                  & $-28746.236^{***}$ & $-70463.727^{***}$ & $-58741.031^{***}$ \\
                                             & $(3386.956)$       & $(2942.538)$       & $(2959.556)$       \\
I(in\_place\_oil\_mill\_sm3/100)             & $0.333^{***}$      & $-0.028$           &                    \\
                                             & $(0.009)$          & $(0.023)$          &                    \\
I(cost\_index/10)                            & $-0.005$           & $-0.019^{***}$     & $-0.000$           \\
                                             & $(0.006)$          & $(0.006)$          & $(0.004)$          \\
year                                         & $28.793^{***}$     & $70.938^{***}$     & $58.352^{***}$     \\
                                             & $(3.393)$          & $(2.951)$          & $(2.955)$          \\
I(year^2)                                    & $-0.007^{***}$     & $-0.018^{***}$     & $-0.015^{***}$     \\
                                             & $(0.001)$          & $(0.001)$          & $(0.001)$          \\
I(price/10)                                  & $0.030$            & $0.026$            & $0.008$            \\
                                             & $(0.015)$          & $(0.015)$          & $(0.010)$          \\
I(price\_l1/10)                              & $0.050^{**}$       & $0.034^{*}$        & $0.020$            \\
                                             & $(0.016)$          & $(0.015)$          & $(0.010)$          \\
I(price\_l2/10)                              & $0.070^{***}$      & $0.058^{***}$      & $0.055^{***}$      \\
                                             & $(0.017)$          & $(0.015)$          & $(0.010)$          \\
I(price\_l3)                                 & $0.004^{*}$        &                    &                    \\
                                             & $(0.002)$          &                    &                    \\
I(price\_l4)                                 & $0.003$            &                    &                    \\
                                             & $(0.002)$          &                    &                    \\
I(price\_l5/10)                              & $-0.008$           & $0.010$            & $-0.003$           \\
                                             & $(0.013)$          & $(0.012)$          & $(0.008)$          \\
I(price\_l6/10)                              & $-0.001$           & $0.007$            & $-0.003$           \\
                                             & $(0.012)$          & $(0.011)$          & $(0.008)$          \\
I(price\_l7/10)                              & $0.009$            & $0.008$            & $0.002$            \\
                                             & $(0.012)$          & $(0.011)$          & $(0.008)$          \\
I(price\_l8/10)                              & $-0.004$           & $-0.001$           & $-0.021^{***}$     \\
                                             & $(0.010)$          & $(0.009)$          & $(0.006)$          \\
EDF: s(prod\_year)                           & $6.505^{***}$      &                    &                    \\
                                             & $(7.326)$          &                    &                    \\
I(price\_l3/10)                              &                    & $0.024$            & $0.031^{**}$       \\
                                             &                    & $(0.015)$          & $(0.010)$          \\
I(price\_l4/10)                              &                    & $0.046^{***}$      & $0.026^{**}$       \\
                                             &                    & $(0.013)$          & $(0.009)$          \\
EDF: s(prod\_year,name)                      &                    & $33.287^{***}$     &                    \\
                                             &                    & $(72.000)$         &                    \\
EDF: s(prod\_year,in\_place\_oil\_mill\_sm3) &                    &                    & $28.195^{***}$     \\
                                             &                    &                    & $(28.469)$         \\
\hline
AIC                                          &                    &                    &                    \\
BIC                                          &                    &                    &                    \\
Log Likelihood                               &                    &                    &                    \\
Deviance                                     & 2391636.349        & 1980765.291        & 914018.219         \\
